MCC Models

### 2019-06-28 papers/full_paper/SI_appendix.Rmd
refactor supplemental model and worked example

@@

The pragmatics model captures the following process: A pragmatic listener ($L_1$) reasons about a pragmatic speaker ($S_1$) who is reasoning about a literal listener ($L_0$) who interprets utterances according to their literal semantics. The outcome is the posterior probability of a particular referent given an utterance. According to Bayes rule it is defined as:

\begin{equation} 
P_{L_1}(r,L|u)\propto P_{S_1}(u|r)P(L)P(r)
\end{equation}

The right side of this equation has three components, the likelihood $P_{S1}(u|r)$, the probability of the lexicon that is considered ($P(L)$) and the prior $P(r)$. Because we studied word learning, we assumed that the listener ($L_1$) had no prior expectations that a particular lexicon (i.e. word - object mapping) is more likely. In the model, $P(L)$ therefore took the form of a uniform distribution over potential lexical.

The likelihood represent the probability that ($S_1$) uses a particular utterance to refer to the referent. It is defined in terms of the informativity of an utterance for a referent:

@@

The informativity of an utterance for a referent is the (log) probability that a naive listener $L_0$ would select that referent given that utterance: Because the listener is assumed to not know the lexicon, the informativity of the utterance averages over the possible lexical, or words meanings, that $L_0$ considers

\begin{equation} 
Informativity(u; r) = ln P(r|u) \propto ln \sum_L P(r,l|u) P(L) \propto P(r) L_{lit}
\end{equation}

while $P(r)$ again denotes the prior probability of a referent, $L_{lit}$ encodes the literal meaning of an utterance given a particular lexicon, returning 1 if the utterance is literally true and 0 if the utterance is literally false. As mentioned above, because $L_1$ does not know the lexicon, the semantics of the words contained in the utterance offer no information about the referent. The non-linguistic aspects of the utterance, however, do. As described in the main text, the utterance also included the agent's turn to one of the tables. The semantics of turning to one of the tables is roughly equal to using a word that could be used to describe the objects on that table. That is, $L_{lit}$ returns 1 for a referent that is on the table the agent turned to and 0 for referents on the other table.

@@

### Numerical Example

In the following, we walk through a numerical example for how model predictions were generated for the pragmatics model. The prediction will correspond to the parameter free model, excluding the noise parameter. The numbers are taken from the preference - different speaker - congruent condition. Fig. 1 shows a screenshot from the adult experiment, which the model was designed to capture. In this context, there are three potential referents (two red-ish, one yellow-ish), of two types (red-ish and yellow-ish) on two tables. In principle, the speaker, the frog in this case, can perform one of two utterances: $u_1$, which is turning to the right table (two objects) and making a request or $u_2$, which involves turning to the left table (single object) and making a request. In the experiment, the speaker chose $u_1$ and we therefore model $L_1$'s interpretation of this utterance.

From a learning perspective, we assume that the listener is learning the mappings between labels and object types instead of object tokens. That is, the listener either thinks the novel word refers to "red-ish objects" or "yellow-ish objects".

The prior distribution over objects (left to right in Fig.1), based on Experiment 2A, for this condition was [`r round(1-mean(adult_ex2_preference_data%>%filter(condition == "different_speaker")%>%pull(correct)),2)`; `r round(1-mean(adult_ex2_preference_data%>%filter(condition == "different_speaker")%>%pull(correct)),2)`; `r round(mean(adult_ex2_preference_data%>%filter(condition == "different_speaker")%>%pull(correct)),2)`]. That is, the yellow object has a higher a priory probability compared to red objects. We start with calculating the informativity of of the utterance for each referent (red or yellow object) based on equation 3.

$$
\begin{aligned}
&Informativity(u_1; r_1) = log(0.64 \cdot 1) = -0.44 \\
&Informativity(u_1; r_2) = log(0.36 \cdot 1) = -1.03
\end{aligned}
$$

with $r_1$ being the yellow object and $r_2$ being the red object. Next we can insert these values into equation 2 to compute the probability that the speaker chooses this utterance to refer to this type of object. To do so, we need the speaker optimality parameter of $\alpha = 2.24$, which was estimated based on Experiment 1.

$$
\begin{aligned}
&P_{S_1}(u_1|r_1) = \frac{exp(2.24 \cdot -0.44)}{exp(2.24 \cdot -0.44)} = 1 \\
&P_{S_1}(u_1|r_2) = \frac{exp(2.24 \cdot -1.03)}{exp(2.24 \cdot -1.03) + exp(2.24 \cdot 0)} = 0.09
\end{aligned}
$$

Here, the informativity for a referent is normalized across all utterances that could be used to refer to this referent. Finally, based on equation 1 we can use these values to compute the probability that the speaker thinks that the speaker is referring to the yellow object ($r_1$) when they produce $r_1$:

>>>>

## Ontology

The situation we model is defined by three sets: referents $r$, utterances $u$, and lexica $\mathcal{L}$. 
Referents are defined by two features: the type of object they are (visually discernible given their shape and color) and their location (one of two tables).
There are two types of objects $t_1$, $t_2$ and two locations $l_1$ and $l_2$.
The utterances available to a speaker are (action, label) pairs, where the action involves turning (or pointing) to a particular location (a table) and labels are novel words ($w_1$, $w_2$), which are assumed to refer to the type of the object. 
The action of turning or pointing reduces the set of referents to those that are on the table targeted by the turn (point); interpreting a label reduces that set further to those which are of the type referred to by that label.
The interpretation of a label requires a lexicon $\mathcal{L}$, which provides the label--type mappings.
There are two lexica: one which maps label 1 $w_1$ to type 1 $t_1$ and label 2 $w_2$ to type 2 $t_2$, and one which has the inverse mapping.

@@

Our word-learning model is a model of pragmatic reasoning couched in the Rational Speech Act modeling framework. The model describes the following process: A pragmatic listener ($L_1$) jointly infers a referent (what object is being picked out by the utterance) and a lexicon (label--type mappings) by reasoning about a pragmatic speaker ($S_1$) who produces utterances to convey information to a literal listener ($L_0$), who in turn interprets utterances according to their literal meaning. According to Bayes rule, this inference is given by:

\begin{equation} 
P_{L_1}(r, \mathcal{L}|u)\propto P_{S_1}(u|r)P( \mathcal{L})P(r)
\end{equation}

The right-hand side of this equation has three components: the prior distribution over referents $P(r)$, the prior distribution over lexica $P(\mathcal{L})$, and the likelihood that a speaker would produce the an utterance $u$ given a referent $r$ ($P_{S1}(u|r)$).
In the situation we model, the prior on referents $P(r)$ is a categorical distribution over three objects in a scene, which we posit could be non-uniform due to what is in common ground (see below for common ground manipulations). 
Because the labels produced by the speaker are all novel words, the listener has no substantive knowledge about the lexica (label--type mappings) and the prior over the two lexica (described above) is uniform.

The listener updates their beliefs about both the referent and the lexicon by reasoning about the speaker, assumed to produce utterances to convey the referent to the listener by being a soft-max rational agent (with degree of rationality $\alpha$) with a utility function defined in terms of the informativity of an utterance for a referent:

@@

The informativity of an utterance for a referent is the (log) probability that a naive listener $L_0$ would select that referent given that utterance.

\begin{equation} 
Informativity(u; r) = \ln P_{L_0}(r|u)
\end{equation}

We assume that the speaker believes the literal listener to also be uncertain about the lexicon; that is, the pragmatic listener believes that the speaker (correctly) believes that the listener does not know the lexicon. 
Because the literal listener is assumed to not know the lexicon, the informativity of the utterance averages over the possible lexical, or words meanings, that $L_0$ considers:

\begin{equation} 
P_{L_0}(r|u) = \sum_L P_{L_0}(r, \mathcal{L}|u)
\end{equation}

where $P_{L_0}(r, \mathcal{L}|u)  \propto \mathcal{L}_{lit} P(r) P(\mathcal{L})$.
$P(r)$ again denotes the prior probability of a referent, $\mathcal{L}_{lit}$ encodes the literal meaning of an utterance given a particular lexicon, returning 1 if the utterance is literally true and 0 if the utterance is literally false. As mentioned above, because $L_1$ does not know the lexicon, the semantics of the words contained in the utterance (i.e., the labels) offer no information about the referent. The non-linguistic aspect of the utterance (the turning or pointing), however, do. As described in above, the semantics of turning to one of the tables is roughly equivalent to saying "It's an object on that table". That is, $\mathcal{L}_{lit}$ returns 1 for a referent that is on the table the agent turned to and 0 for referents on the other table.

### Worked Example

In the following, we work through a numerical example for how model predictions were generated for the pragmatics model. The prediction will correspond to the parameter free model described above (excluding the noise parameter). The values of the parameters are taken from the preference - different speaker - congruent condition. Fig. 1 shows a screenshot from the adult experiment, which the model was designed to capture. In this context, there are three potential referents (two red-ish: $r_1^r$, $r_2^r$; one yellow-ish, $r_3^y$,), of two types (red-ish and yellow-ish) on two tables. 
In principle, the speaker, the frog in this case, can perform one of four utterances, turning either to the left or right table and producing either label ("dax" or "wug"):  $u_1 = (\text{left}, dax)$, $u_2 = (\text{left}, wug)$, $u_3 = (\text{right}, dax)$, $u_4 = (\text{right}, wug)$.
We work out the example where the speaker produces utterance $u_3$, turning to the right table (two objects) and saying the label "dax".

<!-- $u_1$, which is turning to the right table (two objects) and making a request by using a novel label or $u_2$, which involves turning to the left table (single object) and making a request by using a novel label. In the experiment, the speaker chose $u_1$ and we therefore model $L_1$'s interpretation of this utterance. -->

The listener is learning the mappings between labels and object types (rather than object tokens). That is, the listener either thinks the novel word "dax" refers to "red-ish objects" or "yellow-ish objects". Turning or pointing to a table always has the same meaning. These semantics can be described using two lexica: $\mathcal{L_1} = \{dax: \text{red-ish thing}, wug: \text{yellow-ish thing}, point: \text{location of point}\}$, $\mathcal{L_2} = \{dax: \text{yellow-ish thing}, wug: \text{red-ish thing}, point: \text{location of point}\}$.

MH: SHOULD MAKE THIS A DISTRIBUTION (sums to 1).

The prior distribution over referents $P(r)$ (left to right in Fig.1), based on Experiment 2A, for this condition was [`r round(1-mean(adult_ex2_preference_data%>%filter(condition == "different_speaker")%>%pull(correct)),2)`; `r round(1-mean(adult_ex2_preference_data%>%filter(condition == "different_speaker")%>%pull(correct)),2)`; `r round(mean(adult_ex2_preference_data%>%filter(condition == "different_speaker")%>%pull(correct)),2)`]. That is, the yellow object has a higher a priori probability compared to red objects, because the speaker previously displayed a preference for a yellow object.

The literal listener's posterior distribution over referents and lexica a support of size six, owing to the unique combinations of referents and lexica. 
$$
\begin{aligned}
& P_{L_0}(r_1^r, \mathcal{L}_1 | {u_3})  \propto \mathcal{L}_1(r_1^r, u_3) P(r_1^r) P(L_1) = 0 \times 0.36 \times 0.5\\
& P_{L_0}(r_2^r, \mathcal{L}_1 | {u_3})  \propto \mathcal{L}_1(r_2^r, u_3) P(r_2^r) P(L_1) = 1 \times 0.36 \times 0.5\\
& P_{L_0}(r_3^y, \mathcal{L}_1 | {u_3})  \propto \mathcal{L}_1(r_3^y, u_3) P(r_3^y) P(L_1) = 0 \times 0.64 \times 0.5\\
& P_{L_0}(r_1^r, \mathcal{L}_2 | {u_3})  \propto \mathcal{L}_2(r_1^r, u_3) P(r_1^r) P(L_2) = 0 \times 0.36 \times 0.5\\
& P_{L_0}(r_2^r, \mathcal{L}_2 | {u_3})  \propto \mathcal{L}_2(r_2^r, u_3) P(r_2^r) P(L_2) = 0 \times 0.36 \times 0.5\\
& P_{L_0}(r_3^y, \mathcal{L}_2 | {u_3})  \propto \mathcal{L}_2(r_3^y, u_3) P(r_3^y) P(L_2) = 1 \times 0.64 \times 0.5\\
\end{aligned}
$$

because the speaker is pointing to the table with $r_2$ and $r_3$, and under $\mathcal{L_1}$, "dax" means the red object and under $\mathcal{L_2}$, "dax" means the yellow object.
After normalization we have:

$$
\begin{aligned}
& P_{L_0}(r_1^r, \mathcal{L}_1 | {u_3}) = 0 \\
& P_{L_0}(r_2^r, \mathcal{L}_1 | {u_3}) = 0.36 \\
& P_{L_0}(r_3^y, \mathcal{L}_1 | {u_3})   = 0 \\
& P_{L_0}(r_1^r, \mathcal{L}_2 | {u_3})   = 0 \\
& P_{L_0}(r_2^r, \mathcal{L}_2 | {u_3})   = 0 \\
& P_{L_0}(r_3^y, \mathcal{L}_2 | {u_3}) = 0.64\\
\end{aligned}
$$

then,

$$
\begin{aligned}
&P_{L_0}(r_1|u_3) =  P_{L_0}(r_1, \mathcal{L_1}|u_3) + P_{L_0}(r_1, \mathcal{L_2}|u_3) = 0+0=0 \\
&P_{L_0}(r_2|u_3) =  P_{L_0}(r_2, \mathcal{L_1}|u_3) + P_{L_0}(r_2, \mathcal{L_2}|u_3) = 0.36+0=0.36 \\
&P_{L_0}(r_3|u_3) =  P_{L_0}(r_3, \mathcal{L_1}|u_3) + P_{L_0}(r_3, \mathcal{L_2}|u_3) = 0+64=0.64 \\
\end{aligned}
$$

and

$$
\begin{aligned}
&Informativity(u_3; r_1) = log(0) = -\infty \\
&Informativity(u_3; r_2) = log(0.36) = -1.03 \\
&Informativity(u_3; r_3) = log(0.64) = -0.44
\end{aligned}
$$

Next we can insert these values into equation 2 to compute the probability that the speaker chooses this utterance to refer to this type of object. To do so, we need the speaker optimality parameter of $\alpha = 2.24$, which was estimated based on Experiment 1.

SEEMS LIKE WE NEED TO NORMALIZE ACROSS ALL UTTERANCES, SO WE WILL NEED TO REDO the ABOVE CALCULATIONS FOR EACH OF THE 6 UTTERANCES.

$$
\begin{aligned}
&P_{S_1}(u_3|r_2) = \frac{exp(2.24 \cdot -1.03)}{exp(2.24 \cdot -1.03) + exp(2.24 \cdot 0)} = 0.09 \\
&P_{S_1}(u_3|r_3) = \frac{exp(2.24 \cdot -0.44)}{exp(2.24 \cdot -0.44)} = 1 \\
\end{aligned}
$$

Here, the informativity for a referent is normalized across all utterances that could be used to refer to this referent.

Finally, based on equation 1 we can use these values to compute the probability that the speaker thinks that the speaker is referring to the yellow object ($r_1$) when they produce $r_1$:

<!-- The utterances available to a speaker are (action, label) pairs, where the action involves turning (or pointing) to a particular location (i.e., a table) and labels are novel words, which are assumed to refer to the type of the object.  -->
<!-- The action of the turning or pointing rules out the referents that are not on the table targeted by the turn (point).  -->
<!-- The listener knows that the labels refer to the kind of object but does not know which kind is referred to be which label, formalized by the prior over lexica $P(L)$.  -->

<!-- In principle, the speaker, the frog in this case, can perform one of two utterances: $u_1$, which is turning to the right table (two objects) and making a request by using a novel label or $u_2$, which involves turning to the left table (single object) and making a request by using a novel label. -->

<!-- For example, in SI Figure 1, two possible labels are "dax" and "wug", each of which could either mean. -->

<!-- Because we studied word learning, we assumed that the listener ($L_1$) had no prior expectations that a particular lexicon  is more likely. In the model, $P(L)$ therefore took the form of a uniform distribution over potential lexica.  -->

@@
